% ============================================================
\section{Code}
% ============================================================

% ============================================================
\subsection{SQL Patterns (PostgreSQL)}
% ============================================================

Reference for recurring query shapes. Each entry states when to reach for
it and the mechanism that makes it work.

\begin{keybox}[Logical processing order]
A query is evaluated in this order, regardless of written order:
\[
  \texttt{FROM/JOIN} \to \texttt{WHERE} \to \texttt{GROUP BY} \to
  \texttt{HAVING} \to \texttt{SELECT} \to \texttt{DISTINCT} \to
  \texttt{ORDER BY} \to \texttt{LIMIT}.
\]
Two consequences drive most of the patterns below. \texttt{WHERE} runs
before aggregation, so it cannot see aggregates or \texttt{SELECT} aliases.
Window functions are computed at the \texttt{SELECT} step, so they also
cannot appear in \texttt{WHERE} and must be wrapped in a subquery or CTE
to filter on.
\end{keybox}

\paragraph{WHERE vs HAVING.} \texttt{WHERE} filters individual rows before
grouping. \texttt{HAVING} filters whole groups after aggregation and is the
only place an aggregate predicate is legal. A filter on a plain
(non-aggregated) column belongs in \texttt{WHERE} so rows are dropped before
the grouping work happens. Reserve \texttt{HAVING} for conditions like
\texttt{COUNT(*) > 5}.

\vspace{1.5ex}

\begin{lstlisting}[style=sql, caption={WHERE prunes rows, HAVING prunes groups}]
SELECT department, COUNT(*) AS headcount
FROM employees
WHERE active = true          -- row filter, before grouping
GROUP BY department
HAVING COUNT(*) > 5;         -- group filter, on an aggregate
\end{lstlisting}

\vspace{-2ex}

\paragraph{Joins.}
\begin{itemize}[label=$\circ$]
  \item \textbf{INNER:} rows with a match on both sides.
  \item \textbf{LEFT / RIGHT:} keep all rows from one side, \texttt{NULL}-fill the other.
  \item \textbf{FULL OUTER:} keep unmatched rows from both sides.
  \item \textbf{CROSS:} Cartesian product, every left row against every right row.
  \item \textbf{SELF:} a table joined to itself under an alias (employee to manager, row to prior row).
  \item \textbf{Semi-join} (\texttt{EXISTS}, \texttt{IN}): keep left rows with at least one match. No row duplication, no right-side columns.
  \item \textbf{Anti-join} (\texttt{NOT EXISTS}, or \texttt{LEFT JOIN ... WHERE right.key IS NULL}): keep left rows with no match.
\end{itemize}

\paragraph{Rank-then-filter with ROW\_NUMBER().} Use this for top-$N$ per
group when you need to keep the full row. \texttt{GROUP BY} would collapse the
other columns away. Number rows inside each partition, then filter on that
number in an outer query, since the window function cannot live in
\texttt{WHERE}.

Choose the ranker by tie policy: \texttt{ROW\_NUMBER} assigns distinct numbers
and breaks ties arbitrarily, \texttt{RANK} repeats a rank and leaves gaps,
\texttt{DENSE\_RANK} repeats with no gaps.

\begin{lstlisting}[style=sql, caption={Highest-paid employee per department}]
SELECT department, name, salary
FROM (
    SELECT department, name, salary,
           ROW_NUMBER() OVER (PARTITION BY department
                              ORDER BY salary DESC) AS rn
    FROM employees
) ranked
WHERE rn = 1;
\end{lstlisting}

\paragraph{Gaps and islands.} An \emph{island} is a maximal run of consecutive
values. The trick: subtract a dense row number from the value. Across a
consecutive run both sides increase by one per step, so the difference stays
constant, and it jumps exactly at a gap. Group by the difference to collapse
each run.

\begin{lstlisting}[style=sql, caption={Islands of consecutive ids}]
WITH numbered AS (
    -- DISTINCT first so duplicate ids do not desync the row number
    SELECT DISTINCT log_id,
           ROW_NUMBER() OVER (ORDER BY log_id) AS rn
    FROM logs
)
SELECT MIN(log_id) AS island_start,
       MAX(log_id) AS island_end,
       COUNT(*)    AS length
FROM numbered
GROUP BY log_id - rn
ORDER BY island_start;
\end{lstlisting}

The \texttt{DISTINCT} is load-bearing. If \texttt{log\_id} repeats, \texttt{rn}
still advances on the duplicate while the value does not, so
\texttt{log\_id - rn} shifts and splits an island that should stay whole. For
consecutive \emph{dates} rather than integers, offset by an interval:
\texttt{log\_date - rn * INTERVAL `1 day'} is constant within a run.

\begin{lstlisting}[style=sql, caption={Gaps via LEAD}]
SELECT log_id + 1  AS missing_from,
       next_id - 1 AS missing_to
FROM (
    SELECT log_id,
           LEAD(log_id) OVER (ORDER BY log_id) AS next_id
    FROM logs
) t
WHERE next_id - log_id > 1;   -- values missing between the two present ids
\end{lstlisting}

\newpage

\paragraph{Conditional aggregation (pivot).} Fold a categorical column into one
row of counts by wrapping a \texttt{CASE} in an aggregate. Rows the
\texttt{CASE} does not match contribute 0 to \texttt{SUM}. In Postgres the
\texttt{FILTER} clause does the same thing and reads more cleanly.

\begin{lstlisting}[style=sql, caption={Pivot risk categories to columns}]
SELECT
    inspection_type,
    COUNT(*) FILTER (WHERE risk_category IS NULL)           AS no_risk,
    COUNT(*) FILTER (WHERE risk_category = 'Low Risk')      AS low_risk,
    COUNT(*) FILTER (WHERE risk_category = 'Moderate Risk') AS moderate_risk,
    COUNT(*) FILTER (WHERE risk_category = 'High Risk')     AS high_risk,
    COUNT(*)                                                AS total
FROM sf_restaurant_health_violations
GROUP BY inspection_type
ORDER BY total DESC;
\end{lstlisting}

The \texttt{SUM(CASE WHEN ... THEN 1 ELSE 0 END)} form is equivalent and
portable across engines. \texttt{COUNT} ignores \texttt{NULL}, so
\texttt{COUNT(CASE WHEN cond THEN 1 END)} (no \texttt{ELSE}) also counts only
the matches.

\paragraph{Gotchas.}
\begin{itemize}
  \item \textbf{\texttt{NOT IN} with NULLs:} if the subquery returns any
  \texttt{NULL}, \texttt{x NOT IN (...)} is never true and the query returns
  zero rows. \texttt{x NOT IN (1, NULL)} expands to
  \texttt{NOT (x=1 OR x=NULL)}, and the \texttt{NULL} comparison poisons the
  result to unknown. Use \texttt{NOT EXISTS}, which is null-safe.
  \item \textbf{Integer division truncates:} \texttt{5 / 2} yields \texttt{2}
  because the fractional part is discarded. Cast one operand:
  \texttt{5::numeric / 2} or \texttt{5.0 / 2}. Bites hardest in ratio and
  percentage columns.
  \item \textbf{\texttt{COUNT(col)} vs \texttt{COUNT(*)}:} the former skips
  \texttt{NULL}s in \texttt{col}, the latter counts every row. A common source
  of off-by-$N$ errors.
\end{itemize}









%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\newpage


% ============================================================
\subsection{Python Patterns}
% ============================================================

Good solutions to interview problems are often determined by the choice of container and knowledge of which built-in primitives express the required operation.

% ------------------------------------------------------------
\subsubsection*{Containers and cost profiles}

Selecting a container is selecting a cost profile. That decision precedes the
algorithm and frequently dominates the runtime.

\begin{center}
\begin{tabular}{lllll}
\toprule
& \textbf{Ordered} & \textbf{Mutable} & \textbf{Lookup} & \textbf{Use} \\
\midrule
\texttt{list}  & yes & yes & $O(n)$ & ordered sequence, duplicates permitted \\
\texttt{tuple} & yes & no  & $O(n)$ & fixed record, or a dictionary key \\
\texttt{set}   & no  & yes & $O(1)$ & membership, deduplication \\
\texttt{dict}  & insertion & yes & $O(1)$ & mapping from key to value \\
\bottomrule
\end{tabular}
\end{center}

\medskip

\begin{keybox}[Membership cost]
\texttt{x in some\_list} is $O(n)$. \texttt{x in some\_set} is $O(1)$ on
average. A membership test inside a loop over $n$ elements is therefore
$O(n^2)$ against a list and $O(n)$ against a set. Converting the list to a set
once, before the loop, removes a factor of $n$. This substitution accounts for
a large share of naive-to-optimal improvements.
\end{keybox}

\paragraph{Hashability.} Set elements and dictionary keys must be hashable,
which in practice requires immutability. A \texttt{tuple} is a valid key. A
\texttt{list} is not, since mutation would invalidate its position in the hash
table. Converting a list to a tuple to form a composite key is therefore a
standard construction, as in grouping anagrams by their sorted letters.

\begin{lstlisting}[style=python, caption={Tuple as a composite key}]
seen = set()
seen.add((row, col))              # legal, tuple is hashable
# seen.add([row, col])            # TypeError: unhashable type
\end{lstlisting}

\paragraph{Dictionary access.} Counting, grouping, and seen-before problems
reduce to a dictionary. Three access forms differ in their treatment of a
missing key.

\begin{lstlisting}[style=python, caption={Dictionary access}]
d[k]                  # raises KeyError if k is absent
d.get(k, default)     # returns default instead of raising
d.setdefault(k, [])   # inserts default if absent, then returns it

for k in d:                    # iterates over keys
for k, v in d.items():         # iterates over pairs
\end{lstlisting}

% ------------------------------------------------------------
\newpage
\subsubsection*{Built-in functions}

An accumulator loop is often a built-in written out longhand. The following
cover most cases.

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Built-in} & \textbf{Expresses} \\
\midrule
\texttt{len, sum, min, max} & size and aggregation \\
\texttt{sorted} & ordering, parameterized by \texttt{key} and \texttt{reverse} \\
\texttt{any(...)} & existential quantifier: at least one element satisfies $p$ \\
\texttt{all(...)} & universal quantifier: every element satisfies $p$ \\
\texttt{enumerate} & index paired with value \\
\texttt{zip} & parallel iteration over several sequences \\
\texttt{reversed, range} & traversal order \\
\texttt{abs, round, divmod} & numeric utilities \\
\bottomrule
\end{tabular}
\end{center}

\texttt{any} and \texttt{all} short-circuit. The expression
\texttt{any(x < 0 for x in xs)} terminates at the first negative value.

\paragraph{The \texttt{key} argument.} \texttt{sorted}, \texttt{min}, and
\texttt{max} accept a \texttt{key} function applied to each element. Ranking
is performed on the image of that function while the original element is
returned. The consequence is that a record can be ranked on one of its fields
and recovered whole.

\begin{lstlisting}[style=python, caption={Ranking on a field, returning the record}]
min(scores)                            # returns a number
min(records, key=lambda r: r[1])       # returns the full record

sorted(words, key=len)                        # rank by length
sorted(records, key=lambda r: (-r[1], r[0]))  # score desc, name asc
\end{lstlisting}

A tuple key is the analogue of a multi-column \texttt{ORDER BY}. Negating a
single numeric field reverses that field alone. \texttt{reverse=True} cannot
express this, since it inverts every field simultaneously.

% ------------------------------------------------------------
\subsubsection*{Comprehensions}

A comprehension expresses selection and projection in one expression. The form
\texttt{[f(x) for x in xs if p(x)]} corresponds to selecting $f(x)$ over the
elements for which $p(x)$ holds. Four bracket types produce four containers.

\begin{lstlisting}[style=python, caption={The four forms}]
[x*x for x in xs if x > 0]        # list
{x*x for x in xs}                 # set, deduplicates
{k: v for k, v in pairs}          # dict
(x*x for x in xs)      # generator, lazy, builds no intermediate list
\end{lstlisting}

The generator form is appropriate when the result is consumed once, for
example inside \texttt{sum(...)} or \texttt{max(...)}, since it avoids
materializing a list. Nesting beyond two levels is generally less readable
than an explicit loop.

% ------------------------------------------------------------

\newpage
\subsubsection*{Sorting}

\begin{itemize}[itemsep=-0.5ex]
  \item \texttt{sorted(xs)} returns a new list. \texttt{xs.sort()} sorts in
        place and returns \texttt{None}. The assignment \texttt{xs = xs.sort()}
        binds \texttt{None} and fails silently.
  \item Python's sort is \textbf{stable}. Elements comparing equal retain
        their original relative order. Stability licenses sorting in multiple
        passes: sorting by the secondary field first and by the primary field
        second preserves the secondary ordering within each group.
  \item Sorting a sequence of lists or tuples compares elementwise from the
        left. The result depends on field order, so an explicit \texttt{key}
        should be supplied whenever the intent is nonobvious.
  \item Sorting costs $O(n \log n)$, which is negligible at the input sizes
        these problems use.
\end{itemize}

% ------------------------------------------------------------
% ------------------------------------------------------------
\subsubsection*{The \texttt{collections} module}

Four types cover nearly every use. Three are specializations of the
dictionary that eliminate the same recurring boilerplate: the check for
whether a key already exists before writing to it. The fourth replaces the
list when insertion or removal happens at the front.

\medskip

\noindent\textbf{\texttt{Counter}: frequency.}
A \texttt{Counter} is a dictionary from element to count. It is constructed in
one pass over any iterable and answers the question ``how many times does each
value occur.'' Frequency is the most common intermediate object in string and
array problems, and constructing one by hand requires an existence check on
every element.

\begin{lstlisting}[style=python, caption={Counter}]
from collections import Counter

c = Counter("mississippi")  # Counter({'i': 4, 's': 4, 'p': 2, 'm': 1})
c['s']                         # 4
c['z']              # 0, a missing key reads as zero and does NOT raise

c.most_common(2)               # [('i', 4), ('s', 4)], top 2 by count
c.most_common()             # every pair, ordered by count descending
c.total()                      # sum of all counts
\end{lstlisting}

\vspace{-2ex}

A missing key evaluates to \texttt{0} rather than raising \texttt{KeyError},
which is what permits \texttt{c[x] += 1} to be written unconditionally. The
read does not insert the key, so the counter does not grow through inspection.

\texttt{most\_common(k)} selects the top $k$ with a heap in $O(n \log k)$.
Calling \texttt{most\_common()} with no argument sorts the whole counter at
$O(n \log n)$, so the argument should be supplied whenever $k$ is known.

Counters support multiset arithmetic, which expresses several problems
directly.

\begin{lstlisting}[style=python, caption={Counter arithmetic}]
a, b = Counter("aabbc"), Counter("abbbd")

a + b        # counts added
a - b        # counts subtracted, results <= 0 are dropped
a & b        # intersection: elementwise minimum
a | b        # union: elementwise maximum
\end{lstlisting}

\begin{keybox}[Two canonical Counter identities]
\textbf{Anagram test.} Two strings are anagrams if and only if their counters
are equal, so \texttt{Counter(s) == Counter(t)} settles the question in $O(n)$.
Sorting both strings answers the same question in $O(n \log n)$.

\medskip

\textbf{Sliding-window matching.} To locate every permutation of a pattern $p$
inside a text $t$, maintain a counter over a window of length $|p|$ and
compare it against \texttt{Counter(p)} after each shift. Incrementing the
entering character and decrementing the leaving one updates the window in
$O(1)$.
\end{keybox}

\medskip

\noindent\textbf{\texttt{defaultdict}: grouping.}
A \texttt{defaultdict} is constructed with a zero-argument factory. Accessing
a missing key calls that factory, inserts the result, and returns it. The
effect is to remove the initialization branch from every accumulation loop.

\begin{lstlisting}[style=python, caption={defaultdict}]
from collections import defaultdict

groups = defaultdict(list)          # missing key -> []
for name, dept in employees:
    groups[dept].append(name)       # no existence check required

counts = defaultdict(int)           # missing key -> 0
adj    = defaultdict(set)           # missing key -> set(), an adjacency list
grid   = defaultdict(lambda: defaultdict(int))   # a nested, sparse 2D table
\end{lstlisting}

The factory is any callable taking no arguments. \texttt{int()} returns
\texttt{0}, \texttt{list()} returns \texttt{[]}, and \texttt{set()} returns an
empty set, which is why the bare type name is passed rather than a call to it.
A nested default requires a \texttt{lambda}, since
\texttt{defaultdict(defaultdict(int))} would pass an instance where a factory
is expected.

\texttt{defaultdict(list)} performs a \texttt{GROUP BY} that omits the
aggregation step. Every member of each group survives, which is the property
that \texttt{ROW\_NUMBER() OVER (PARTITION BY ...)} supplies in SQL. Building a
graph's adjacency list is the same operation applied to edges.

\begin{keybox}[Reading a \texttt{defaultdict} mutates it]
Any access to a missing key inserts it. The expression \texttt{if d[k]:}
therefore creates \texttt{k} with an empty default and the dictionary grows
through inspection alone. This corrupts a subsequent iteration over
\texttt{d.items()} and inflates \texttt{len(d)}.

For a read that must not write, use \texttt{k in d} or \texttt{d.get(k)}.
Neither triggers the factory.
\end{keybox}

The alternative to a \texttt{defaultdict} is
\texttt{d.setdefault(k, []).append(x)} on an ordinary dictionary. It produces
the same result and constructs a fresh empty list on every call, including the
calls where the key is already present. The \texttt{defaultdict} factory runs
only on a miss.

\medskip

\noindent\textbf{\texttt{deque}: a queue.}
A \texttt{deque} is a double-ended queue backed by a doubly linked list of
fixed-size blocks. Appending and popping at either end is $O(1)$. The relevant
comparison is against a \texttt{list}, which stores its elements contiguously,
so removing the front element requires shifting every remaining element left by
one position.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Operation} & \textbf{\texttt{list}} & \textbf{\texttt{deque}} \\
\midrule
Append or pop at the right & $O(1)$ & $O(1)$ \\
Append or pop at the left  & $O(n)$ & $O(1)$ \\
Index at position $i$      & $O(1)$ & $O(n)$ \\
\bottomrule
\end{tabular}
\end{center}

The costs are complementary. A \texttt{list} is the correct stack, since a
stack only touches the right end. A \texttt{deque} is the correct queue.
Random access into a \texttt{deque} is a sign that the wrong container was
chosen.

\begin{lstlisting}[style=python, caption={deque}]
from collections import deque

q = deque([1, 2, 3])
q.append(4)        # right
q.appendleft(0)    # left
q.pop()            # from the right
q.popleft()        # from the left, O(1) where list.pop(0) is O(n)
q.rotate(1)        # shift every element right by one, cyclically
\end{lstlisting}

Breadth-first search is the standard application. The frontier is processed in
insertion order, which requires removal from the front on every iteration of
the main loop.

\begin{lstlisting}[style=python, caption={BFS frontier}]
q = deque([start])
seen = {start}
while q:
    node = q.popleft()        # FIFO, so nodes are visited by distance
    for nxt in adj[node]:
        if nxt not in seen:
            seen.add(nxt)
            q.append(nxt)
\end{lstlisting}

\paragraph{Bounded length.} A \texttt{deque} constructed with \texttt{maxlen=k}
discards an element from the opposite end whenever a full deque is appended to.
This maintains a rolling window over the last $k$ items with no explicit
eviction logic, which is useful for moving averages and for any statistic over
a fixed lookback.

\begin{lstlisting}[style=python, caption={A rolling window of the last 3 values}]
w = deque(maxlen=3)
for x in stream:
    w.append(x)          # the oldest element is evicted automatically
    if len(w) == 3:
        print(sum(w) / 3)
\end{lstlisting}

\medskip

\noindent\textbf{\texttt{namedtuple}: a labeled record.}
A tuple carries no information about what its positions mean, so \texttt{r[1]}
requires the reader to remember the field order. A \texttt{namedtuple} attaches
names while remaining a tuple, so it stays immutable, hashable, unpackable, and
usable as a dictionary key or a set element.

\begin{lstlisting}[style=python, caption={namedtuple}]
from collections import namedtuple

Student = namedtuple("Student", ["name", "score"])
s = Student("Ada", 92.0)

s.score            # 92.0, and s[1] still works
name, score = s    # unpacking is unchanged
min(students, key=lambda s: s.score)     # the key function reads clearly
\end{lstlisting}

The cost over a plain tuple is negligible and the gain is that a \texttt{key}
function or a sort becomes self-documenting. When mutability is required,
\texttt{dataclasses.dataclass} provides the equivalent with assignable fields.

% ------------------------------------------------------------
\subsubsection*{Iteration}

The following forms eliminate manual index arithmetic, which is the usual
source of off-by-one errors.

\begin{lstlisting}[style=python, caption={Iteration forms}]
for i, x in enumerate(xs):        # index and value together
for a, b in zip(xs, ys):          # parallel, halts at the shorter sequence
for a, b in zip(xs, xs[1:]):      # consecutive pairs
for x in reversed(xs):            # reverse traversal without index arithmetic

name, score = record              # tuple unpacking
first, *rest = xs                 # starred unpacking
\end{lstlisting}

\vspace{-2ex}

The construction \texttt{zip(xs, xs[1:])} yields every adjacent pair and
underlies run detection, difference sequences, and monotonicity checks. It is
the analogue of \texttt{LAG} in SQL.

% ------------------------------------------------------------
\subsubsection*{Problem structure}

\begin{keybox}[The four-move skeleton]
A typical warm-up problem decomposes into four steps.
\begin{enumerate}
  \item \textbf{Parse} the input into a list of records.
  \item \textbf{Reduce} the records to a target key: a minimum, a maximum, a
        count, or the $k$-th distinct value.
  \item \textbf{Filter} the records against that key.
  \item \textbf{Order and format} the output.
\end{enumerate}
Step 2 accounts for most errors. The characteristic failure is fusing steps 2
and 3 into a single pass that mishandles an edge case.
\end{keybox}

\paragraph{Ranking and ties.} The most frequent error in step 2 concerns
ties. Whether tied elements share a rank or consume one is the same decision
in Python as in SQL.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Tie policy} & \textbf{Python} & \textbf{SQL} \\
\midrule
Ties consume a rank & \texttt{sorted(xs)[k-1]}       & \texttt{ROW\_NUMBER() = k} \\
Ties share a rank   & \texttt{sorted(set(xs))[k-1]}  & \texttt{DENSE\_RANK() = k} \\
\bottomrule
\end{tabular}
\end{center}

For scores \texttt{[20, 20, 30]}, the second lowest is \texttt{30}. Both
elements equal to \texttt{20} hold rank 1 and neither occupies rank 2, so
\texttt{sorted(xs)[1]} evaluates to \texttt{20} and is incorrect. The
\texttt{set} converts \texttt{ROW\_NUMBER} semantics into \texttt{DENSE\_RANK}
semantics.

\paragraph{Rank then filter.} Once the key is identified, it is computed in
one pass and the records are filtered against it in a second. The structure
mirrors the necessity of wrapping a window function in a CTE, since a window
function cannot appear in \texttt{WHERE}.

\begin{lstlisting}[style=python, caption={Rank then filter}]
target  = sorted({score for _, score in records})[1]   # rank
winners = sorted(name for name, score in records if score == target)
                                            # filter, then order
\end{lstlisting}

% ------------------------------------------------------------
\subsubsection*{Common errors}

\begin{itemize}
  \item \texttt{sorted(set(xs))[1]} raises \texttt{IndexError} when all values
        are identical.
  \item Slicing copies. \texttt{xs[a:b]} costs $O(b-a)$, so a slice inside a
        loop makes the loop quadratic.
  \item \texttt{list.insert(0, x)} and \texttt{list.pop(0)} are $O(n)$, since
        every subsequent element shifts. A \texttt{deque} provides $O(1)$
        operations at both ends.
  \item Integer division \texttt{//} floors toward negative infinity, so
        \texttt{-7 // 2} evaluates to \texttt{-4}. \texttt{int(-7 / 2)}
        truncates toward zero.
  \item A mutable default argument (\texttt{def f(xs=[])}) is created once at
        definition time and persists across calls. The default should be
        \texttt{None}, with the container constructed inside the function body.
  \item Strings are immutable. Accumulating with \texttt{s += c} in a loop is
        $O(n^2)$. Collect the pieces in a list and apply
        \texttt{''.join(parts)}.
\end{itemize}

% ============================================================
\subsection{Algorithmic Paradigms}
% ============================================================

Each paradigm is identified by a recognition cue in the problem statement.

\begin{itemize}
  \item \textbf{Hashing.} Cue: find a pair, count occurrences, seen before,
        contains a duplicate. Space is traded for $O(1)$ lookup. The canonical
        construction tests whether the complement of the current element
        already appears in a map of previously visited elements.

  \item \textbf{Two pointers.} Cue: a sorted sequence, or a pairing problem.
        Pointers converging from both ends solve ``two elements summing to
        $t$'' in $O(n)$ with constant extra space. Fast and slow pointers
        detect a cycle in a linked list.

  \item \textbf{Sliding window.} Cue: a contiguous subarray or substring
        subject to a constraint. The right edge expands to admit elements and
        the left edge contracts while the constraint is violated. Each element
        enters and leaves once, giving $O(n)$.

  \item \textbf{Prefix sums.} Cue: repeated range-sum queries over a static
        array. Cumulative sums are precomputed once, after which any range
        query is a difference of two entries. This is the array analogue of
        \texttt{SUM() OVER}.

  \item \textbf{Binary search on the answer.} Cue: the smallest $x$ such that
        a condition is feasible. Requires a monotone predicate, meaning that
        feasibility at $x$ implies feasibility at every larger $x$. The search
        ranges over the answer space rather than the input.

  \item \textbf{Sorting as preprocessing.} Cue: intervals, scheduling, or
        grouping equal elements. Sorting converts a global problem into a local
        one in which each element need only be compared against its neighbor.
\end{itemize}

\begin{keybox}[Complexity anchors]
\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Operation} & \textbf{Cost} \\
\midrule
\texttt{x in set}, \texttt{x in dict}   & $O(1)$ average \\
\texttt{x in list}                      & $O(n)$ \\
\texttt{list.append}                    & $O(1)$ amortized \\
\texttt{list.insert(0,x)}, \texttt{list.pop(0)} & $O(n)$ \\
\texttt{deque.appendleft}, \texttt{deque.popleft} & $O(1)$ \\
\texttt{sorted(xs)}                     & $O(n \log n)$ \\
\texttt{xs[a:b]}                        & $O(b-a)$, copies \\
\bottomrule
\end{tabular}
\end{center}
\end{keybox}

% ============================================================
\newpage
\subsection{Recursion and Memoization}
% ============================================================

A recursive function is defined by three obligations. Omitting any one of them
produces either a wrong answer or a stack overflow.

\begin{keybox}[The recursion contract]
\begin{enumerate}
  \item \textbf{Base case.} At least one input for which the answer is
        returned without a recursive call.
  \item \textbf{Recursive case.} The answer for input $n$ expressed in terms
        of the answers for strictly smaller inputs.
  \item \textbf{Progress.} Every recursive call moves the argument closer to a
        base case, so the recursion terminates.
\end{enumerate}
Correctness follows by induction. Assuming the recursive call returns the
correct answer on the smaller input, verify only that the combination step is
correct. Tracing the full call tree by hand is unnecessary and is the main
reason recursion feels harder than it is.
\end{keybox}

\paragraph{The call stack.} Each pending call occupies a stack frame, so the
space cost of a recursion is $O(d)$ where $d$ is the maximum depth. Python
imposes a default recursion limit of 1000 frames and raises
\texttt{RecursionError} beyond it. A recursion whose depth scales with $n$ is
therefore unsafe for large $n$ unless it is rewritten as a loop with an
explicit stack.

\paragraph{Overlapping subproblems.} Naive recursion recomputes identical
subproblems. The Fibonacci recurrence $F(n) = F(n-1) + F(n-2)$ generates a
call tree with $O(\varphi^n)$ nodes, since $F(n-2)$ is evaluated once by the
$F(n-1)$ branch and again directly. Caching each result by its argument
collapses the tree to $O(n)$ distinct evaluations.

\begin{lstlisting}[style=python, caption={Memoization by decorator}]
from functools import lru_cache

@lru_cache(maxsize=None)          # or @cache in Python 3.9+
def fib(n):
    if n < 2:                     # base case
        return n
    return fib(n-1) + fib(n-2)    # recursive case
\end{lstlisting}

Arguments to a memoized function must be hashable, which is the reason
memoized state is passed as an integer, a string, or a tuple rather than as a
list or a set.

\paragraph{Top-down and bottom-up.} Memoized recursion is the top-down form.
It computes only the states that are reachable, at the cost of a call stack.
Tabulation is the bottom-up form. It fills an array in an order that
guarantees every dependency is already resolved, using no stack and no
recursion.

\begin{lstlisting}[style=python, caption={The same recurrence, tabulated}]
def fib(n):
    if n < 2:
        return n
    prev, curr = 0, 1
    for _ in range(n - 1):
        prev, curr = curr, prev + curr    # only two states are ever live
    return curr
\end{lstlisting}

The final line illustrates a general reduction. When a recurrence depends only
on the previous $k$ states, the table collapses from $O(n)$ space to $O(k)$.

% ------------------------------------------------------------
\subsubsection*{Dynamic programming}

Dynamic programming applies when a problem has \textbf{optimal substructure},
meaning the optimal solution is composed of optimal solutions to subproblems,
together with \textbf{overlapping subproblems}, meaning those subproblems recur.
Absent overlap, plain divide and conquer suffices. Absent optimal substructure,
a greedy or exhaustive method is required.

\begin{keybox}[Formulating a DP]
Five questions, answered in order, before any code is written.
\begin{enumerate}
  \item \textbf{State.} What arguments fully determine a subproblem? This is
        the signature of the function or the index set of the table.
  \item \textbf{Recurrence.} How is a state's value expressed in terms of
        smaller states?
  \item \textbf{Base cases.} Which states are known outright?
  \item \textbf{Order.} In what order can states be evaluated so that
        dependencies are resolved first? Top-down recursion answers this
        automatically.
  \item \textbf{Answer.} Which state holds the final result?
\end{enumerate}
The complexity is the number of states multiplied by the cost of evaluating
one state.
\end{keybox}

\paragraph{Canonical state definitions.} Most problems reuse a small number of
state shapes.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{State} & \textbf{Meaning} & \textbf{Typical problem} \\
\midrule
\texttt{dp[i]}    & best over the prefix ending at $i$ & maximum subarray, house robber \\
\texttt{dp[i][j]} & best over two prefixes             & edit distance, longest common subsequence \\
\texttt{dp[i][w]} & first $i$ items, capacity $w$      & 0/1 knapsack \\
\texttt{dp[i][j]} & best over the interval $[i,j]$     & matrix chain, palindrome partition \\
\bottomrule
\end{tabular}
\end{center}

% ------------------------------------------------------------
\subsubsection*{Backtracking}

Backtracking is recursion over a search tree with an explicit undo step. It
enumerates permutations, subsets, and combinations, and prunes branches that
cannot yield a valid solution.

\begin{lstlisting}[style=python, caption={Backtracking skeleton}]
def solve(path, choices, results):
    if is_complete(path):
        results.append(path[:])        # copy, since path is mutated in place
        return
    for c in choices:
        if not is_valid(path, c):
            continue                   # prune
        path.append(c)                 # choose
        solve(path, remaining(choices, c), results)
        path.pop()                     # undo
\end{lstlisting}

The \texttt{path[:]} copy is mandatory. Appending \texttt{path} itself stores a
reference to a list that subsequent mutation will alter, so every recorded
result ends up identical.

For unconstrained enumeration, the standard library is preferable to a
hand-written recursion.

\begin{lstlisting}[style=python, caption={Enumeration without recursion}]
from itertools import permutations, combinations, product

permutations(xs, r)     # ordered arrangements of length r
combinations(xs, r)     # unordered selections of length r
product(xs, repeat=r)   # r-tuples with repetition, the r-fold Cartesian power
\end{lstlisting}

These correspond exactly to the counting formulas $P(n,r) = n!/(n-r)!$,
$\binom{n}{r}$, and $n^r$.

% ============================================================
\subsection{Heaps and Priority Queues}
% ============================================================

A binary heap maintains the minimum of a collection under insertion and
deletion at logarithmic cost. Python's \texttt{heapq} operates on an ordinary
list interpreted as a heap, so no dedicated type is involved.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Operation} & \textbf{Call} & \textbf{Cost} \\
\midrule
Convert a list in place & \texttt{heapify(xs)}     & $O(n)$ \\
Insert                  & \texttt{heappush(h, x)}  & $O(\log n)$ \\
Remove the minimum      & \texttt{heappop(h)}      & $O(\log n)$ \\
Inspect the minimum     & \texttt{h[0]}            & $O(1)$ \\
Push then pop           & \texttt{heappushpop(h, x)} & $O(\log n)$, one pass \\
\bottomrule
\end{tabular}
\end{center}

A heap provides no ordering beyond its root. Iterating over the underlying
list does not produce sorted output.

\paragraph{Minimum only.} \texttt{heapq} implements a min-heap exclusively. A
max-heap is obtained by negating the keys on insertion and negating again on
extraction. For records, push the tuple \texttt{(-priority, item)}, since
tuples compare lexicographically and the first field dominates.

\begin{keybox}[Top-$k$ under a stream]
To retain the $k$ largest of $n$ elements, maintain a min-heap of size $k$.
Each new element is compared against the root, which is the smallest of the
current $k$. If the element is larger, the root is evicted and the element
inserted.

The cost is $O(n \log k)$ and the memory is $O(k)$. Sorting the full input
costs $O(n \log n)$ and requires all $n$ elements in memory at once, so the
heap is superior when $k \ll n$ and is the only option when the input arrives
as a stream of unknown length.
\end{keybox}

\begin{lstlisting}[style=python, caption={The k largest elements}]
import heapq

def top_k(stream, k):
    h = []                              # min-heap holding the k best so far
    for x in stream:
        if len(h) < k:
            heapq.heappush(h, x)
        elif x > h[0]:                  # h[0] is the weakest survivor
            heapq.heapreplace(h, x)     # pop the root, then push x
    return sorted(h, reverse=True)
\end{lstlisting}

When the whole input already fits in memory, the library provides this
directly as \texttt{heapq.nlargest(k, xs)} and \texttt{heapq.nsmallest(k, xs)},
both of which accept a \texttt{key} argument.

\paragraph{Two heaps for a running median.} Maintain a max-heap over the lower
half of the values and a min-heap over the upper half, rebalancing after each
insertion so their sizes differ by at most one. The median is then read from
the root of the larger heap, or is the mean of the two roots when the sizes
are equal. Insertion costs $O(\log n)$ and the median query is $O(1)$.

\todo{\textbf{TODO:} Add graph traversal (BFS, DFS, and the
\texttt{deque}-based queue) alongside the heap entry, since Dijkstra composes
the two. Add a string-processing section covering \texttt{split},
\texttt{join}, and sliding windows over characters.}